Differential invariant signatures for planar Lie group transformations of images

### 2020-08-13 paper/newpaper.tex

\subsection{$SE(2)$}

\todo[inline]{Make precise!}
Ignoring translation, the action of $SE(2)$ on the first derivatives is simply rotation by some angle $\theta$, i.e.,

\begin{equation}
\left( \begin{array}{c} f_x \\ f_y \end{array} \right) = \left( \begin{array}{cc} \cos \theta & \sin \theta \\ - \sin \theta & \cos \theta \end{array} \right) \left( \begin{array}{c} \bar f_{\bar x} \\ \bar f_{\bar y} \end{array} \right).
\end{equation}

Using $c$ for $\cos \theta$ and $s$ for $\sin \theta$ and prolonging the group to second order via the chain rule gives:

\begin{equation}
\arraycolsep=1.4pt
\left(
\begin{array}{c}
  \bar f_x \\ \bar f_y \\ \bar f_{xx} \\ \bar f_{xy} \\ \bar f_{yy} 
 \end{array}
 \right)
 = 
\left(
\begin{array}{ccccc}
 c & -s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & cs & c^2 - s^2 & -cs \\
0 & 0 & s^2 & 2cs & c^2 \\
\end{array}
\right)\left(
\begin{array}{c}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
 \end{array}
 \right).
\label{eqn:SE2}
 \end{equation}

As $SE(2)$ does not include scaling, lengths stay the same, and so an appropriate choice for first-order normalisation is $\bar f_x = \sqrt(f_x^2 + f_y^2),\:\: \bar f_y = 0$, which sets $c = f_x / \| \nabla f \|$, $s = - f_y / \| \nabla f \|$. This provides the invariant $I_1 = \nabla f$. As there are no more free parameters, the expressions for the second-order terms can be read off as:

\begin{align}
I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} \\
I_3 &= f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2) \\
I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} 
\end{align}

\todo[inline]{ Two things to note: (1) SE(2) has similar invars, (2) there is an issue in E(2) with the moving frame: reflection ambiguity. That's why we did it by contraction.
}

>>>>

\subsection{The Special Euclidean Group $SE(2)$}
An element of the special Euclidean group $SE(2)$ acts on an element $(x, y)$ of $\mathbb{R}^2$ to produce $(\bar{x}, \bar{y})$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &= x\cos\theta  - y\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
We prolong the group action to form the jet space $J^2$ with coordinates $(x,
y, f, f_x, f_y, f_{xx}, f_{xy}, f_{yy})$ via the chain rule. The
transformation rule for the derivatives is given by
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix} = 
  \begin{bmatrix}
 c & -s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & cs & c^2 - s^2 & -cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix},
\label{eqn:SE2}
 \end{equation}
where $c = \cos \theta$ and $s = \sin \theta$. Because the group has three
parameters, we choose the cross section of the group orbits to be $\bar{x}
= 0, \bar{y} = 0, \bar{f}_{\bar{y}} = 0$. We also require that $\bar{f}_{x}
> 0$ to uniquely define the moving frame. Note that the group action is
locally free (away from critical points) when prolonged to $J^1$, however
we prolonged to $J^2$ in order to compute invariants. points. The moving
frame is then given by the prolonged Euclidean transformation that maps an
element of  $J^2$ to this cross section. It can be thought of as rotating
the image about that point so that its gradient is pointing in the positive
$x$ direction, and then translating the image so the point is mapped to the
origin.

This choice of moving frame gives $\cos\theta = f_x/\lVert\nabla f\rVert$
and $\sin\theta = -f_y/\lVert \nabla f \rVert$. The parameters $t_x, t_y$
are then $t_x = -x\cos\theta + y\sin\theta$ and $t_y = -x\sin\theta -
y\cos\theta$, however these have no effect on the derivatives, and will
henceforth be ignored.

The remaining elements not fixed by the cross section, $\bar{f}$,
$\bar{f}_{\bar{x}}$, $\bar{f}_{\bar{x}\bar{x}}$,
$\bar{f}_{\bar{x}\bar{y}}$, and $\bar{f}_{\bar{y}\bar{y}}$ are therefore
all invariant. This gives the following set of invariants:
\begin{equation}\label{eq:se2invariants}
\begin{aligned}
  I_0 &= f \\
  I_1 &= f_x^2 + f_y^2 & &(= \bar{f}_{\bar{x}}^2)\\
  I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} &&(= I_1\bar{f}_{\bar{x}\bar{x}}) \\
  I_3 &= f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2) &&(= I_1\bar{f}_{\bar{x}\bar{y}})\\
  I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} &&(= I_1\bar{f}_{\bar{y}\bar{y}})
\end{aligned}
\end{equation}
Note that we multiplied each of the second derivative invariants by $I_1$
so that they would decay smoothly to zero at critical points.

\todo[inline]{c.f. transvectant formulae}

\subsection{The Euclidean Group E(2)}
An element of the Euclidean group $E(2)$ maps an element $(x, y)$ of
$\mathbb{R}^2$ to $(\bar{x}, \bar{y})$ as 
\begin{equation*}
  \begin{aligned}
    \bar{x} &=  x\varepsilon\cos\theta  - y\varepsilon\sin\theta + t_x \\
    \bar{y} &= x\sin\theta  + y\cos\theta + t_y.
  \end{aligned}
\end{equation*}
where $\varepsilon \in \{-1, 1\}$ and the remaining parameters are the same
as for $SE(2)$. If $\varepsilon = 1$ it is a rigid transformation, and if
$\varepsilon = -1$ it contains a reflection.

This time, the group action is not free on $J^1$ (with elements of the form
$(x, y, f, f_x, f_y)$) away from critical points because there are two
possible transformations that map to the cross section $(0, 0, \bar{f},
\bar{f}_{\bar{x}}, 0)$ where $\bar{f}_{\bar{x}} > 0$, one that reflects the
gradient into the positive $x$ direction, and one that rotates.

We again prolong the group action to $J^2$, but this time we expect to need
to use a second derivative term to determine $\varepsilon$ for the moving
frame. The derivative transformations are given by
\begin{equation}
  \begin{bmatrix}
  \bar f_{\bar x} \\ \bar f_{\bar y} \\ \bar f_{\bar{x}\bar{x}} \\ \bar f_{\bar{x}\bar{y}} \\ \bar f_{\bar{y}\bar{y}}
  \end{bmatrix} = 
  \begin{bmatrix}
 \varepsilon c & -\varepsilon s & 0 & 0 & 0 \\
 s & c & 0 & 0 & 0 \\
0 & 0 & c^2 & -2cs & s^2 \\
0 & 0 & \varepsilon cs & \varepsilon(c^2 - s^2) & -\varepsilon cs \\
0 & 0 & s^2 & 2cs & c^2 \\
  \end{bmatrix}
  \begin{bmatrix}
f_x \\ f_y \\ f_{xx} \\ f_{xy} \\ f_{yy} 
  \end{bmatrix},
\label{eqn:E2prolongation}
\end{equation}
where again $c = \cos\theta$ and $s = \sin\theta$.

We begin by choosing the same cross section as for $SE(2)$, namely $\bar{x} = 0,
\bar{y} = 0, \bar{f}_{\bar{y}} = 0, \bar{f}_{\bar{x}} > 0$. This gives
$\cos\theta = \varepsilon f_x / \lVert \nabla f \rVert$ and $\sin\theta =
-\varepsilon f_y / \lVert \nabla f \rVert$. Under this transformation,
regardless of whether $\varepsilon = 1$ or $\varepsilon = -1$, the
derivative $\bar{f}_{\bar{x}} = \lVert \nabla f \rVert$. We need a second
derivative term to resolve the choice of $\varepsilon$ for the moving
frame. The only second derivative term in which $\varepsilon$ appears is
the $\bar{f}_{\bar{x}\bar{y}}$ term. Substituting in for $\cos\theta,
\sin\theta$, we see that
\begin{equation*}
    \bar{f}_{\bar{x}\bar{y}} = \frac{\varepsilon(f_xf_y(f_{yy} - f_{xx}) + (f_x^2 -
    f_y^2)f_{xy})}{f_x^2 + f_y^2}
\end{equation*}
This suggests choosing $\varepsilon$ to make $\bar{f}_{\bar{x}\bar{y}} \ge
0$. We can then list the invariants:
\begin{equation}\label{eq:e2invariants}
\begin{aligned}
  I_0 &= f \\
  I_1 &= f_x^2 + f_y^2 & &(= \bar{f}_{\bar{x}}^2)\\
  I_2 &= f_x^2 f_{xx} + 2 f_x f_y f_{xy} + f_y^2 f_{yy} &&(= I_1\bar{f}_{\bar{x}\bar{x}}) \\
  I_3 &= \lvert f_x f_y (f_{yy} - f_{xx}) + f_{xy} (f_x^2 - f_y^2)\rvert
  &&(= I_1\lvert \bar{f}_{\bar{x}\bar{y}}\rvert)\\
  I_4 &= f_y^2 f_{xx} - 2 f_x f_y f_{xy} + f_x^2 f_{yy} &&(= I_1\bar{f}_{\bar{y}\bar{y}})
\end{aligned}
\end{equation}
Notice that most, but not all of the $SE(2)$ invariants
\eqref{eq:se2invariants} are also $E(2)$ invariants. We again multiply
through the invariants by $I_1$ so that they decay smoothly at critical
points.
